TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2018-09-21 paper/draft.tex
rm old comments
add ref

% HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before). Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract. Right now it just feels a bit hacked-together, since this is basically just my AAS abstract that we kept adding some sentences to as we went, and I think it needs a better frame. I'm still trying to get a sense of what is scientifically interesting about this sample, and I've been struggling to pin point what the real relevance of this work is: it's not exactly just that the Cannon is precise for M dwarfs (this original scope, like an extension of Ness, is too boring I think, given what we've learned), and it's not really about M dwarf physics or why we are better than models (this is too beyond us to argue).

% I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure. Is this what we want to emphasize?

% Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add? It seems like just because we have \gaia\ parallaxes we should go ahead and do it? Maybe also to do a full-circle comparison with our test sample labels.

@@

In particular, parameters such as metallicity can be difficult to determine precisely, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star (under the assumption kinematically similar pairs are co-evolutionary), to developed precise empirical relations (as good as $\sim$0.07dex) between the precise adopted metallicities of the FGK primary, and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis. \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

% HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise. 
% For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc). However these metallicity callibrations are inaccesible for APOGEE spectra, because the wavelength coverage does not include the relevant callibration indices. \thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.

>>>>

@@

In particular, parameters such as metallicity can be difficult to determine precisely, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis. \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
